---
title: "lab_02"
author: "derek willis"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most?
**A1.** 26 departments have offered courses with "Climate" in the title. Atmospheric and Oceanic Science has the most classes offered, 68. 

```{r}
climate_classes_seats1 <- umd_courses |> filter(str_detect(title,"Climate"))|>filter(seats>=1)|>
  group_by(department)|>
  summarise(count_classes =n())

```

**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why?
**A2.** Personally, the courses that are most interesting to me are the Germanic Studies classes, because I took German here at the University of Maryland. I'm simply curious what the classes are. But I'm sure they would be boring. 


**Q3** Which departments have the most courses in the Fall 2023 term? And which departments in that term have at least one syllabus posted? How are the top five departments different in each result?

**A3.** The departments with the most courses offered in the Fall 2023 term are: Business and Management (164), School of Music (137), English (120), Teaching and Learning, Policy and Leadership (116), and Information Studies (115). 124 departments in that term have at least one syllabus posted. Business and Management has the most courses in the Fall AND it is the department that has the most syllabi posted. The five departments that are offering the most classes are listed above. When you filter to see which of the departments has syllabi posted, English and the Teaching Leadership departments foll off the top five list. Looks like these departments weren't as good at posting syllabi as they are at offering classes in the fall of 2023. Economics and Biological Sciences are departments that move up on the list when you search for departments with posted syllabi. It seems these departments are more consistent at posting syllabi for the courses they offer in Fall 2023. 

```{r}
fall_courses_syllabus <- umd_courses |> filter(term ==202308) |>group_by(department)|>
  summarise(count_classes=n())|> 
  arrange(desc(count_classes))
```
```{r}
fall_courses_more_syllabi <- 
  umd_courses|> filter (term ==202308)|>group_by(department)|>
  filter(syllabus_count>=1)|>
  summarise(count_classes=n())|>
  arrange(desc(count_classes))
```


**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as the same dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name). Write a sentence describing the pattern of spending over time.
**A4.** one pattern I notice is that the campaign appears to be spending the most money as it got closer to election date. August, September, October, and November of 2022 account for the top four months of spending. 

```{r}
maryland_expenses <- maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))

head(maryland_expenses)
```
```{r}
maryland_expenses |>
  mutate(month = floor_date(expenditure_date, "month")) |>
  group_by(month) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
```
**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for the others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).
**A5** 77,723,146 dollars were spent in state. 33,164,994	dollars were spent out of state. 

```{r}
maryland_expenses_spending_location<- maryland_expenses |>
  mutate(
    spending_location = case_when(str_detect(address, " Maryland ") ~ "In-state",
        str_detect(address, " California ") ~ "Out of state",
        str_detect(address, " Washington ") ~ "Out of state",
        str_detect(address, " Louisiana ") ~ "Out of state",
        str_detect(address, " Florida ") ~ "Out of state",
        str_detect(address, " North Carolina ") ~ "Out of state",
        str_detect(address, " Massachusetts ") ~ "Out of state",
        str_detect(address, " West Virginia ") ~ "Out of state",
        str_detect(address, " Virginia ") ~ "Out of state",
        .default = NA
      )
  )
```

```{r}
maryland_expenses_spending_location|>
  group_by(spending_location)|>
  summarise(total=sum(amount))|>
  arrange(desc(total))

```

**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab.
**A6.** Using the umd courses data if you filter out all the NA's for professor, what is the average number of classes taught by professors in fall 2023? 

note -I thought I had this figured out, but I didn't. Stepped away for a breath of fresh air. I'm taking the loss on this one, but will keep on working to learn more.  


```{r}
remove_na<-umd_courses|>filter(term==202308)|>
  summarise(count_courses=n())|>
  arrange(desc(count_courses))
 

```
