---
title: "lab_03"
author: "Derek Willis"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)
# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

```{r}
cleaned_loan_data <- loan_data |> clean_names()
cleaned_loan_data

```

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows

agency_payment <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv")

# Load in the Maryland grants and loan data
loan_data <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")
```

```{r}
cleaned_agency_payment <- agency_payment |> clean_names()
cleaned_agency_payment
```
## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? Look at & search the records in the dataframe while you construct your code.

***A1***: I found 5 different possible versions of PG County School System entities and 30 different rows. 

```{r}
#this code should find the variations of Prince George's
# save this result to pg_schools

pg_schools <-cleaned_agency_payment |>
  filter(
    str_detect(vendor_name,"PRINCE GEOREGES COMMUNITUY COLLEGE") |
      str_detect(vendor_name, "PRINCE GEORGE'S COMMUNITY COLLEGE") |
      str_detect(vendor_name,"PRINCE GEORGE PUBLIC SCHOOLS") |
      str_detect(vendor_name,"PRINCE GEORGES COMM COLLEGE") |
      str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL"))



```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you? 

***A2***:FY2014 had the highest amount of payments to the prince george's county school system. It stands out to me that there are several years in which PG county schools did not receive any payment


```{r}
pg_schools |> 
  group_by (fiscal_year) |> 
  summarize(amount = sum(amount)) |> 
  arrange(desc(amount))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?


***A3*** You would need to make sure duplicate vendors are grouped and cross reference those vendors by zip code. Three examples of vendor names that are similar are: Whiting Turner Contracting, Johns Hopkins, and Pocomoke Marketing Partnerships, Inc. I think, to clean the data so that any nearly identical vendors would look identical, I would browse the data to find all the variations of the spelling of one company. Then I would decide on one standardized version of the spelling. I would use mutate and then and "if/then" clause, to transform any spelling variations of the comopany to the standardized version.

```{r}
capital_projects<-
  cleaned_agency_payment |>
  filter(str_detect(agency_name,"CAPITAL PROJECT."))
```

```{R}
capital_projects|>
group_by(vendor_name)|>
  summarise(amount =sum(amount))|>
  arrange(desc(amount))
  
```
Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

***A4***: I can't stay stem cell research in Maryland has decreased over time, but the research appears to have been the highest in the first year of the data set (2009), and it has not increased since then. 

```{r}
cleaned_loan_data_with_scresearch <-
  cleaned_loan_data|>
 mutate(
   sc_research = case_when(
   str_detect(grantor, "stem cell") ~ "sc_research",
     str_detect(grantor, "STEM CELL")~ "sc_research",
     str_detect(grantor, "Stem Cell")~ "sc_research",
   .default = NA
 ))

```

```{r}
cleaned_loan_data_with_scresearch |>
  filter(
    str_detect(sc_research, "sc_research")
  )|>
  group_by(fiscal_year)|>
  summarize(total = sum(amount)) |>
  arrange(desc(total))

```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you? 

***A5*** The code resulted in 18 rows of grantees, however some of the data is messy. One grantee is University of Maryland. Another is University of Maryland College Park. I would want to arrange the grantees in alphabetical order to maximize the chances that I can easily visualize possible duplicate grantees. I also would like to have the data either in all caps or all lowercase. For me the grantees that are NOT the UMD grantees are most interesting for me. It seems prety standard that a major university like UMD would be receiving state grants, but who are these other companies? What affiliation do they have (if any) with the university, and are they hiring? :-) 

```{r}
cleaned_loan_data|>
  filter(str_detect(zip_code, "20742"))|>
  group_by(grantee)|>
  summarise(total =sum(amount))|>
  arrange(desc(total))

          
```

