---
title: "week7_recap"
author: "Daniel Trielli"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Look at the earthquakes dataset. Finish the sentence below as if you were explaining the data to someone who had not seen it before but needs to know about it.

**A1:** This dataset contains data on earthquakes (over the past month) collected by the United States Geological Survey. It shows the depth of earthquakes around the world, their magnitude, when they happened and other interesting facts about earthquakes. 

------------------------------------------------------------------------

#### **Q2** How many records there are there in this dataset? What do they mean and what useful information we can gather from it, looking at the columns?

**A2:**
I see 9775 records in the data (I'm counting the rows). Each record is an earthquake. The useful information we can gather from the earthquake include where the earthquake happened and how powerful of a quake it was. 

------------------------------------------------------------------------

#### **Q3** How do I reorganize this data to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}
#reorganizing the data to show greatest depth first
earthquakes|>
  arrange(desc(depth))

```

**A3:**
The deepest earthquake I found was roughly 670 kilometers deep. 
------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}
##filtering for magnitude larger than 6 and counting
earthquakes|>
  filter(mag>6)|>
  summarise(count = n())
```

**A4:**
There are 13 earthquakes in the data that have a magnitude greater than 6
------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}
#counting earthquakes that meet both criteria
earthquakes|>
  filter(mag>6)|>
  filter(depth<20)|>
  summarise(count =n())

```

**A5:**
There are six earthquakes in the data set that meet both criteria: magnitude larger than 6 and depth smaller than 20 km.
------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 OR a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}
#filtering and counting earthquakes that fit either category
earthquakes|>
  filter(mag>6 |depth<20)|>
  summarise(count = n())

```

**A6:**

I see 7445 earthquakes that are either greater than 6 in magnitude or less that 20 km deep. 

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}
#Filtering to find the # of earthquakes that took place in Alaska
selected_place <- earthquakes|> select(place)
selected_place|>
  filter(str_detect(place, "Alaska"))|>
  summarise(count_earthquakes = n())

            
```

**A7:**
To find the number of earthquakes in Alaska, you should look at the place column. You can use the string detect function to find all of the rows in the place column with the name Alaska in it. There are 3446 earthquakes that took place in Alaska during this period. 
------------------------------------------------------------------------

#### **Q8** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}
earthquakes|>
  group_by(type)|>
summarise(count_earthquakes = n())|>
arrange(desc(count_earthquakes))

```

**A8:**
The best way to find all the types of earthquakes is to group by types and then you count out the number of types of earthquake that occured. The two most common types of quake are earthquakes (9529) and quarry blast (133). 
------------------------------------------------------------------------

#### **Q9** What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}
#finding average depth 
earthquakes|>
  group_by(type)|>
  summarise(count_earthquakes = n(),
  mean_depth = mean(depth))

```

**A9:**
The average depth of earthquakes is 25.25 km deep. I find the negative depth surprising, but I guess it shouldn't be.  

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}
 hourly_earthquakes <- 
  earthquakes|>
  mutate (hour = hour(time))
     
```

**A10:**
You extract the hour by using the mutate function for hour as shown in the text block above. 

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column showing whether an earthquake is serious or not?

```{r}
hourly_earthquakes|>
  mutate(
    serious = case_when(
      mag >3 ~ "yes", 
    .default = "no"
    ))
    

```

**A11:**
You create a new column by using the mutate function, and you label that new column serious (or something to that extent) Then you make sure that any entry where the magnitude is greater than 3, is correctly marked in the serious column. 

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}
hourly_earthquakes|>
  mutate(
    serious = case_when(
      mag >3 ~ "yes", 
    .default = "no"
    ))|>
  group_by(hour,serious)|>
  summarise(count_earthquakes = n())|>
  arrange(count_earthquakes)
  
```

```{r}
hourly_earthquakes|>
  mutate(
    serious = case_when(
      mag >3 ~ "yes", 
    .default = "no"
    ))|>
  group_by(hour,serious)|>
  summarise(count_earthquakes = n())|>
  arrange(desc(count_earthquakes))
  

```

**A12**:

The hour with the least serious earthquakes is twelve noon. The hour with the most serious earthquakes is 2 am with 67 serious earthquakes occuring during this hour.  

------------------------------------------------------------------------

#### **Q13** What's another question you are interested in and how would you ask it in R? Which hour in the data set has the most non serious earthquakes? 

```{r}

hourly_earthquakes|>
  mutate(
    serious = case_when(
      mag >3 ~ "yes", 
    .default = "no"
    ))|>
  group_by(hour,serious)|>
  summarise(count_earthquakes = n())|>
  arrange(desc(count_earthquakes))
  
```
###Answer: The hour with the most non-serious earthquakes is 11 am. 